### 2024-01-18 outputs/paper/paper.qmd
updated paper and updated data cleaning

---
title: "Is Crime Really On A Rise In Toronto? An Analysis of The Past And Current Crime Rates of Toronto"
author: 
  - Adrian Ly
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "Current mainstream news outlets are pushing a narrative that crime in Toronto is on a rise. "
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(knitr)
library(janitor)
library(lubridate)
library(opendatatoronto)
library(tidyverse)
library(dplyr)
```

# Introduction

You can and should cross-reference sections and sub-sections.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

The data utilized throughout this paper was obtained through the City of Toronto's

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("outputs/data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

$$
\text{Percentage Change} = \left( \frac{\text{New Value} - \text{Old Value}}{\text{Old Value}} \right) \times 100
$$

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```
We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @fig-avgcrimerates.

```{r}
#| label: fig-avgcrimerates
#| fig-cap: Average Crime Rates
#| echo: false

# Reading the data

average_crime_rates_per_year <- read_csv("outputs/data/average_crime_rates_data.csv")

# Creating the line graph for average crime rates
ggplot(average_crime_rates_per_year, aes(x = year, y = average_rate, color = crime)) +
  geom_line() +
  labs(title = "Average Crime Rates by Type Across Years (2014-2023)",
       x = "Year",
       y = "Average Crime Rate",
       color = "Crime Type") +
  theme_minimal()
```

```{r}
#| label: fig-percentagecrimerates
#| fig-cap: Average Crime Rates
#| echo: false

# Reading the data
percentage_change_data <- read_csv("outputs/data/percentage_change_data.csv")

# Creating the line graph for percentage changes
ggplot(percentage_change_data, aes(x = year, y = percentage_change, color = crime)) +
  geom_line() +
  labs(title = "Yearly Percentage Change in Average Crime Rates by Type (2014-2023)",
       x = "Year",
       y = "Percentage Change in Average Rate",
       color = "Crime Type") +
  theme_minimal()
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-percentagecrimerates20142023
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

# Reading the data
final_percentage_change_table <- read.csv("outputs/data/percentage_change_table_2014_2023.csv")
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-percentagecrimerates20222023
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

percentage_change_table <- read.csv("outputs/data/percentage_change_table_2022_2023.csv")
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {.unnumbered}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-01-18 scripts/02-data_cleaning.R

@@ -70,5 +70,50 @@ average_crime_rates_per_year <- long_data |>
   group_by(crime, year) |>
   summarize(average_rate = mean(rate, na.rm = TRUE))
 
+# Calculating the percentage change in the average rate per year for each crime type
+percentage_change_data <- average_crime_rates_per_year |>
+  group_by(crime) |>
+  # We want to compare the current avg rate with the previous avg rate to get
+  # our rate change
+  mutate(percentage_change = (average_rate / lag(average_rate) - 1) * 100) |>
+  na.omit() # Remove any NA values that may occur
+
+# Calculating the average crime rate for each crime type in the first and last years
+first_year <- min(long_data$year)
+last_year <- max(long_data$year)
+
+average_rates_first_last_years <- long_data |>
+  filter(year == first_year | year == last_year) |>
+  group_by(crime, year) |>
+  summarize(average_rate = mean(rate, na.rm = TRUE)) |>
+  spread(year, average_rate)
+
+# Calculating the percentage change from first year to last year for each crime type
+final_percentage_change <- average_rates_first_last_years |>
+  mutate(percentage_change = ((`last_year` - `first_year`)/`first_year`) * 100)
+
+# Selecting only relevant columns for the table
+final_percentage_change_table <- final_percentage_change |>
+  select(crime, percentage_change)
+
+# Calculating the average crime rate for each crime type in 2022 and 2023
+average_rates_2022_2023 <- long_data %>%
+  filter(year == 2022 | year == 2023) %>%
+  group_by(crime, year) %>%
+  summarize(average_rate = mean(rate, na.rm = TRUE)) %>%
+  spread(year, average_rate)
+
+# Calculating the percentage change from 2022 to 2023 for each crime type
+percentage_change_2022_2023 <- average_rates_2022_2023 %>%
+  mutate(percentage_change = ((`2023` - `2022`)/`2022`) * 100)
+
+# Selecting only relevant columns for the table
+percentage_change_table <- percentage_change_2022_2023 %>%
+  select(crime, percentage_change)
+
 #### Save data ####
-write_csv(average_crime_rates_per_year, "outputs/data/analysis_data.csv")
+write_csv(average_crime_rates_per_year, "outputs/data/average_crime_rates_data.csv")
+write_csv(percentage_change_data, "outputs/data/percentage_change_data.csv")
+write.csv(final_percentage_change_table, "outputs/data/percentage_change_table_2014_2023.csv")
[…]

### 2024-01-19 outputs/paper/paper.qmd
updated paper

@@
 
-```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
-#| echo: false
-
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
-```
-
 Talk more about it.
@@
 (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
-```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
-#| echo: false
-#| warning: false
-#| message: false
-
-analysis_data <- read_csv(here::here("outputs/data/analysis_data.csv"))
-
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
-  theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
-```
 
@@
 
-average_crime_rates_per_year <- read_csv("outputs/data/average_crime_rates_data.csv")
+average_crime_rates_per_year <- read_csv(here::here("outputs/data/average_crime_rates_data.csv"))
 
@@
 # Reading the data
-percentage_change_data <- read_csv("outputs/data/percentage_change_data.csv")
+percentage_change_data <- read_csv(here::here("outputs/data/percentage_change_data.csv"))
 
@@
 # Reading the data
-final_percentage_change_table <- read.csv("outputs/data/percentage_change_table_2014_2023.csv")
+final_percentage_change_table <- read.csv(here::here("outputs/data/percentage_change_table_2014_2023.csv"))
 ```
@@
 
-percentage_change_table <- read.csv("outputs/data/percentage_change_table_2022_2023.csv")
+percentage_change_table <- read.csv(here::here("outputs/data/percentage_change_table_2022_2023.csv"))
 ```
@@
 
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
-```
-
 ## Diagnostics
@@
 
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
-```
-
 \newpage

### 2024-01-19 outputs/paper/paper.qmd
updated readme and paper

@@
   - Adrian Ly
-thanks: "Code and data are available at: LINK."
+thanks: "Code and data analysis are available at: https://github.com/AdrianUofT/Toronto-Crime-Rates-Investigation.git"
 date: today
 date-format: long
-abstract: "Current mainstream news outlets are pushing a narrative that crime in Toronto is on a rise. "
+abstract: "Current mainstream news outlets are pushing a narrative that crime in Toronto is on a rise.
+Overall crime rates when looking from a macro scale perspective does seem to be increasing on average of about 0.43%, with the micro scale showcasing some crimes such as homicide to have an increase of almost 180%.
+Using data from OpenDataToronto, this paper serves to present an unbiased review of the current crime rates in Toronto and to reduce any fearmongering that may occur when someone reads a headline.
+This paper reveals that while crime is technically on a rise, this \"increase\" in crime rate does not necessairly correlate to something worth pointing out."
 format: pdf
@@
 #| echo: false
+#| message: false
 
@@
 final_percentage_change_table <- read.csv(here::here("outputs/data/percentage_change_table_2014_2023.csv"))
+
+final_percentage_change_table |> kable()
 ```
@@
 percentage_change_table <- read.csv(here::here("outputs/data/percentage_change_table_2022_2023.csv"))
+
+percentage_change_table |> kable()
 ```
@@
 
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
+ is a trace plot. It shows... This suggests...
 
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
+ is a Rhat plot. It shows... This suggests...
 

### 2024-01-19 outputs/paper/paper.qmd
updated paper

@@
 
-The data utilized throughout this paper was obtained through the City of Toronto's
-
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
-
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
-Talk way more about it.
+The data utilized throughout this paper was obtained through the City of Toronto's OpenDataToronto [@citeOpenDataToronto].
+The specific data set that was used is entitled 'Neighbourhood Crime Rates, 2023' [@citeData].
+Data was collected and analyzed through the statistical program 'R' [@citeR]
 
@@
 ```
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
+We run the model in R [@citeR] using the `rstanarm` package of . We use the default priors from `rstanarm`.
 
@@
 
-final_percentage_change_table |> kable()
+colnames(final_percentage_change_table) <- c("X", "Crime", "Percentage Change")
+
+final_percentage_change_table |> select("Crime", "Percentage Change") |> kable()
 ```
@@
 
-percentage_change_table |> kable()
+colnames(percentage_change_table) <- c("X", "Crime", "Percentage Change")
+percentage_change_table |> select("Crime", "Percentage Change") |> kable()
 ```
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
+In we implement a posterior predictive check. This shows...
 
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
+In we compare the posterior with the prior. This shows...
 

### 2024-01-19 scripts/02-data_cleaning.R

@@ -97,18 +97,18 @@ final_percentage_change_table <- final_percentage_change |>
   select(crime, percentage_change)
 
 # Calculating the average crime rate for each crime type in 2022 and 2023
-average_rates_2022_2023 <- long_data %>%
-  filter(year == 2022 | year == 2023) %>%
-  group_by(crime, year) %>%
-  summarize(average_rate = mean(rate, na.rm = TRUE)) %>%
+average_rates_2022_2023 <- long_data |>
+  filter(year == 2022 | year == 2023) |>
+  group_by(crime, year) |>
+  summarize(average_rate = mean(rate, na.rm = TRUE)) |>
   spread(year, average_rate)
 
 # Calculating the percentage change from 2022 to 2023 for each crime type
-percentage_change_2022_2023 <- average_rates_2022_2023 %>%
+percentage_change_2022_2023 <- average_rates_2022_2023 |>
   mutate(percentage_change = ((`2023` - `2022`)/`2022`) * 100)
 
 # Selecting only relevant columns for the table
-percentage_change_table <- percentage_change_2022_2023 %>%
+percentage_change_table <- percentage_change_2022_2023 |>
   select(crime, percentage_change)
 
 #### Save data ####

### 2024-01-21 outputs/paper/paper.qmd
updated paper

--- 2024-01-19 outputs/paper/paper.qmd
@@
   - Adrian Ly
-thanks: "Code and data are available at: LINK."
+thanks: "Code and data analysis are available at: https://github.com/AdrianUofT/Toronto-Crime-Rates-Investigation.git"
 date: today
 date-format: long
-abstract: "Current mainstream news outlets are pushing a narrative that crime in Toronto is on a rise. "
+abstract: "Current mainstream news outlets are reporting a significant increase in crime rates, a trend that from our data analysis supports.
+Our findings indicate a dramatic increase in crimes such as homicide and autotheft, with them increasing over 210% from 2014.
+Comparing from 2022, homicide rates have escalted over 180% with autotheft having an increase of 20%.
+These substanistal increases in crime rate underscore a evergrowing concern around public safety in Toronto and highlight the urgent need for effective crime prevention strategies."
 format: pdf
@@
 
+#### Work space set-up ####
 library(knitr)
@@
 library(dplyr)
+
+# Reading in data
+cleaned_crime_rates_data <- read_csv(here::here("outputs/data/cleaned_crime_rates_data.csv"))
+average_crime_rates_per_year <- read_csv(here::here("outputs/data/average_crime_rates_data.csv"))
+percentage_change_data <- read_csv(here::here("outputs/data/percentage_change_data.csv"))
+final_percentage_change_table <- read.csv(here::here("outputs/data/percentage_change_table_2014_2023.csv"))
+percentage_change_table <- read.csv(here::here("outputs/data/percentage_change_table_2022_2023.csv"))
 ```
@@
 
-The data utilized throughout this paper was obtained through the City of Toronto's
+The data utilized throughout this paper was obtained through the City of Toronto's `opendatatoronto` [@citeOpenDataToronto].
+The specific data set that was used is entitled `Neighbourhood Crime Rates, 2023` [@citeData].
+Data was collected and analyzed through the statistical program `R` [@citeR], with additional support from `tidyverse` [@citeTidyverse], `ggplot2` [@citeGGPlot2], `knitr` [@citeKnitR], `janitor` [@citeJanitor], `lubridate` [@citeLubridate], `dplyr` [@citeDplyr], `tibble` [@citeTibble], `readr` [@citeReadR], `stringr` [@citeStringR], and `kableExtra` [@citeKable].
+Further discussion of data colelction, cleaning, and analysis process can be found later in this paper.
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+## Neighbourhood Crime Rate Statistics {#sec-data2}
 
-Talk more about it.
+This data set was published by the Toronto Police Services.
+It outlines the various different crime data by neighbourhood.
+That is, for each neighbourhood, the data set shows the associated area name, its hood ID, the population for said area, followed by the type of crime and number of times said crime has been committed for a specific year.
+The main crimes being analyzed here are assault, auto theft, bike theft, break and enter, robbery, theft over, theft from motor vehicle, homicide, and shooting & firearm discharges.
+The data set also includes the crime rate per 100,000 population.
+That is, it calculates the crime count per 100,000 population per year.
+It is to be noted that for some areas, some crime data may be listed as "NA".
 
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
+Any other types of crime are not represented in this data set and are deemed to be outside the scope of this paper.
+This data set was last freshed on January 14, 2024 and was captured for this paper on January 16, 2024.
+I conducted the first step of cleaning the data to omit any NA values, simplifying the names of columns, eliminating any additional columns and re-factoring the data (see @tbl-crimedata).
 
-Talk way more about it.
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-crimedata
+#| tbl-cap: "Sample of Cleaned Neighbourhood Crime Rate Statistics Data"
+#| warning: false
 
-# Model
+head(cleaned_crime_rates_data) |> kable(col.names = c("Crime", "Year", "Crime Rate"))
+```
 
-The goal of our modelling strategy is twofold. Firstly,...
+## Average Crime Rate
 
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+Using the cleaned crime rate data set, I created a script using the `group_by` function which first allowed me to first group the data by type of crime and year.
+Then using the `summarize` function, I was able to calculate the average the average crime rate for each crime per year (see @tbl-avgcrimedata).
 
-## Model set-up
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-avgcrimedata
+#| tbl-cap: "Sample of Average Crime Rate Statistics Data"
+#| warning: false
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
-
-$$
-\text{Percentage Change} = \left( \frac{\text{New Value} - \text{Old Value}}{\text{Old Value}} \right) \times 100
-$$
-
-```{=tex}
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
+head(average_crime_rates_per_year) |> kable(col.names = c("Crime", "Year", "Average Crime Rate"))
 ```
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
-
-### Model justification
-
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
 
@@
 
-Our results are summarized in @fig-avgcrimerates.
+## Average Yearly Crime Rate
+
+Our results of the average crime rate for all of the neighborhoods in Toronto, from 2014-2023, for each type of crime is summarized in @fig-avgcrimerates.
+As of January 14, 2024, out of the 158 different neighborhoods, the crime with the highest average crime rate is assault with an average crime rate of 806 crimes per 100,000 population.
+The lowest being homicides with an average of 7.6 crimes per 100,000 population.
 
@@
 #| label: fig-avgcrimerates
-#| fig-cap: Average Crime Rates
+#| fig-cap: Average Crime Rates by Type Across Years (2014-2023)
 #| echo: false
-
-# Reading the data
-
-average_crime_rates_per_year <- read_csv(here::here("outputs/data/average_crime_rates_data.csv"))
+#| message: false
 
@@
 
+## Percentage Change in Average Crime Rates
+
+Further breaking down this data allows us to focus on the trends of crime rates over a specified period and determining the percentage change in the average crime rate to give us an insight into how crime patterns have evolved.
+
+### Calcuating the Percentage Change in Average Crime Rate
+
+Given our data set, we computed the percentage change of the average yearly crime rate and analyzed the year-over-year percentage change in these rates.
+The percentage change was computed using the following formula:
+
+$$
+\text{Percentage Change} = \left( \frac{\text{New Value} - \text{Old Value}}{\text{Old Value}} \right) \times 100
+$$
+
+Using the `mutate` function, I was able to calculate the percentage change on a year-over-year basis of the average yearly crime rate for each different type of crime.
+@tbl-percentagecrimerates20142023 shows the percentage change for each crime when comparing the years 2014 and 2023.
+Notably, we can see that both auto theft and homicide has increased over 210% while crimes such as robbery and theft from motor vehicle have both decreased ~24% and ~21% respectively.
+@tbl-percentagecrimerates20222023 shows how homicide is now up 180% while robbery is now up 11% with theft from motor vehicle only down 7%.
+For a full picture, @fig-percentagecrimerates details the percentage changes on a year-over-year basis.
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-percentagecrimerates20142023
+#| tbl-cap: "Percentage Changes in Average Yearly Crime Rate from 2014 to 2023"
+#| warning: false
+
+colnames(final_percentage_change_table) <- c("X", "Crime", "Percentage Change")
+
+final_percentage_change_table |> select("Crime", "Percentage Change") |> kable()
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-percentagecrimerates20222023
+#| tbl-cap: "Percentage Changes in Average Yearly Crime Rate from 2022 to 2023"
+#| warning: false
+
+colnames(percentage_change_table) <- c("X", "Crime", "Percentage Change")
+percentage_change_table |> select("Crime", "Percentage Change") |> kable()
+```
+
 ```{r}
 #| label: fig-percentagecrimerates
-#| fig-cap: Average Crime Rates
+#| fig-cap: Yearly Percentage Change in Average Crime Rates by Type (2014-2023)
 #| echo: false
-
-# Reading the data
-percentage_change_data <- read_csv(here::here("outputs/data/percentage_change_data.csv"))
+#| message: false
 
@@
   geom_line() +
-  labs(title = "Yearly Percentage Change in Average Crime Rates by Type (2014-2023)",
+  labs(title = str_wrap("Yearly Percentage Change in Average Crime Rates by Type (2014-2023)", width = 40),
        x = "Year",
@@
 
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-percentagecrimerates20142023
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
-# Reading the data
-final_percentage_change_table <- read.csv(here::here("outputs/data/percentage_change_table_2014_2023.csv"))
-```
-
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-percentagecrimerates20222023
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
-percentage_change_table <- read.csv(here::here("outputs/data/percentage_change_table_2022_2023.csv"))
-```
-
 # Discussion
 
-## First discussion point {#sec-first-point}
+Overall, the data reveals that there is a significant upward trend of crime rates, more particularly in homicide and auto theft rates.
+The data indicates an over 210% increase in homicide rates with a more notable spike of 180% in the last year.
+Similarly, we have auto theft rising by 20% from last year.
+If we were to factor in all of the crime rates in our data set, the overall crime rate increase in the last year is about ~32%.
+These findings are in alignment with recent reports from various news outlets that suggest a growing trend in criminal activity.
+The [National Post](https://nationalpost.com/news/canada/violent-crime-rose-again-in-2022-murder-rate-highest-in-30-years-statcan) stating 'murder rate highest in 30 years' [@citearticle3], [CP24](https://www.cp24.com/news/toronto-saw-15-per-cent-increase-in-its-violent-crime-index-in-2022-more-than-three-times-the-jump-seen-nationally-statscan-1.6498269?cache=yesclipId10406200text%2Fhtml%3Bcharset%3Dutf-80404%2F7.258454%2F7.632243%2F7.243626%2F7.243626%2F7.243626) claiming that 'Toronto saw 15 per cent increase in its violent crime index in 2022' [@citearticle1], or the [CBC](https://www.cbc.ca/news/canada/toronto/auto-theft-stories-toronto-2023-1.7060776#:~:text=In%20Toronto%20proper%2C%20vehicle%20thefts,jump%20in%202022%20from%202021.) saying that 'vehicle thefts were up some 25.4 per cent over 2022' [@citearticle2].
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+As to the reason why crime rates are on the rise, there isn't necessarily a clear answer.
+What would someone's motive be for committing a crime in the first place?
+This could be down to numerous different factors.
+Some factors that could be contributing to this trend could be socio-economic factors.
+For example income inequality, unemployment, high costs of living, etc.
+which are major prevailing problems that all Canadians, not just specifically people living in Toronto, in the last few years has experienced to some degree.
+Another factor could be the societal attitude towards crime.
+If someone were to commit a crime, there is obviously an incentive to do so.
+In the case of auto theft, thieves can get away with stealing an entire car and then selling it over-seas.
+The point to highlight here is they can get away with doing said crime.
+If the incentive to commit the crime far out-weighs any potential punishment for said crime, then it gives criminals the encouragement to do the crime and more importantly repeat doing so.
+This increase in crime rate calls for an urgent review of our current crime prevention and law enforcement strategies.
 
-## Second discussion point
+There are several limitations to this analysis 
+# Conclusion
 
-## Third discussion point
-
-## Weaknesses and next steps
-
-Weaknesses and next steps should also be included.
-
-\newpage
-
-\appendix
-
-# Appendix {.unnumbered}
-
-# Additional data details
-
-# Model details {#sec-model-details}
-
-## Posterior predictive check
-
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
-
-## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
-
-\newpage
+This paper investigated the
 

### 2024-01-21 scripts/02-data_cleaning.R

@@ -79,18 +79,15 @@ percentage_change_data <- average_crime_rates_per_year |>
   na.omit() # Remove any NA values that may occur
 
 # Calculating the average crime rate for each crime type in the first and last years
-first_year <- min(long_data$year)
-last_year <- max(long_data$year)
-
 average_rates_first_last_years <- long_data |>
-  filter(year == first_year | year == last_year) |>
+  filter(year == 2014 | year == 2023) |>
   group_by(crime, year) |>
   summarize(average_rate = mean(rate, na.rm = TRUE)) |>
   spread(year, average_rate)
 
 # Calculating the percentage change from first year to last year for each crime type
 final_percentage_change <- average_rates_first_last_years |>
-  mutate(percentage_change = ((`last_year` - `first_year`)/`first_year`) * 100)
+  mutate(percentage_change = ((`2023` - `2014`)/`2014`) * 100)
 
 # Selecting only relevant columns for the table
 final_percentage_change_table <- final_percentage_change |>
@@ -112,6 +109,7 @@ percentage_change_table <- percentage_change_2022_2023 |>
   select(crime, percentage_change)
 
 #### Save data ####
+write_csv(long_data, "outputs/data/cleaned_crime_rates_data.csv")
 write_csv(average_crime_rates_per_year, "outputs/data/average_crime_rates_data.csv")
 write_csv(percentage_change_data, "outputs/data/percentage_change_data.csv")
 write.csv(final_percentage_change_table, "outputs/data/percentage_change_table_2014_2023.csv")

### 2024-01-22 outputs/paper/paper.qmd
updated paper
updated paper, readme and LLM usuage

@@
 abstract: "Current mainstream news outlets are reporting a significant increase in crime rates, a trend that from our data analysis supports.
-Our findings indicate a dramatic increase in crimes such as homicide and autotheft, with them increasing over 210% from 2014.
-Comparing from 2022, homicide rates have escalted over 180% with autotheft having an increase of 20%.
-These substanistal increases in crime rate underscore a evergrowing concern around public safety in Toronto and highlight the urgent need for effective crime prevention strategies."
+Our findings indicate a dramatic increase in crimes such as homicide and auto theft, with them increasing over 210% from 2014.
+Compared to 2022, homicide rates have escalated by over 180% with auto theft having an increase of 20%.
+These substantial increases in crime rate underscore an evergrowing concern around public safety in Toronto and highlight the urgent need for effective crime prevention strategies."
 format: pdf
@@
 bibliography: references.bib
+toc: true
 ---
@@
 
-You can and should cross-reference sections and sub-sections.
-
-The remainder of this paper is structured as follows. @sec-data....
+Over the recent years, major news outlets have been increasingly focused on the trends and patterns of crime rates in our city, Toronto.
+The narrative of a major increase in overall crime is sparking debate and concern over the general public's safety.
+This has caused a comprehensive analysis to understand the nature of these trends.
+To dissect and interpret crime data, this paper focuses on various types of crime to provide an objective view of how crime patterns and trends have evolved.
+More specifically, our analysis focuses on more petty thefts such as bike theft to more serious crimes such as homicide and auto theft.
+We examine the data from 2014 to 2023 seeking to uncover the underlying trends and percentage increases or decreases in crime rates.
+The goal is to offer a clear, data-driven perspective beyond anecdotal evidence and media narratives.
+Understanding these trends is not only important to inform the general public about them but also to point the direction for law enforcement and policymakers to make imperative choices to combat these criminals.
+This paper is structured into the following sections: Data, Results, Discussion, and Conclusion.
+In the data section I discuss the nature of these crime rates obtained from the City of Toronto's `opendatatoronto` [@citeOpenDataToronto] library, and the steps I took to clean and analyze the data.
+The result section highlights the trends found in the analysis process.
+The discussion section provides further evaluation of the trends and serves to provide insight as to the potential reasoning behind these crime trends and possible areas that require more attention.
+Lastly, the conclusion section is to summarize the main findings of this paper.
 
@@
 The specific data set that was used is entitled `Neighbourhood Crime Rates, 2023` [@citeData].
-Data was collected and analyzed through the statistical program `R` [@citeR], with additional support from `tidyverse` [@citeTidyverse], `ggplot2` [@citeGGPlot2], `knitr` [@citeKnitR], `janitor` [@citeJanitor], `lubridate` [@citeLubridate], `dplyr` [@citeDplyr], `tibble` [@citeTibble], `readr` [@citeReadR], `stringr` [@citeStringR], and `kableExtra` [@citeKable].
-Further discussion of data colelction, cleaning, and analysis process can be found later in this paper.
+Data was collected and analyzed through the statistical program `R` [@citeR], with additional support from `tidyverse` [@citeTidyverse], `ggplot2` [@citeGGPlot2], `knitr` [@citeKnitR], `janitor` [@citeJanitor], `lubridate` [@citeLubridate], `dplyr` [@citeDplyr], `tibble` [@citeTibble], `readr` [@citeReadR], `stringr` [@citeStringR], `here` [@citeHere], and `kableExtra` [@citeKable].
+Further discussion of data collection, cleaning, and analysis process can be found later in this paper.
 
@@
 This data set was published by the Toronto Police Services.
-It outlines the various different crime data by neighbourhood.
-That is, for each neighbourhood, the data set shows the associated area name, its hood ID, the population for said area, followed by the type of crime and number of times said crime has been committed for a specific year.
-The main crimes being analyzed here are assault, auto theft, bike theft, break and enter, robbery, theft over, theft from motor vehicle, homicide, and shooting & firearm discharges.
+It outlines the various crime data by neighborhood.
+That is, for each neighborhood, the data set shows the associated area name, its hood ID, the population for said area, followed by the type of crime and number of times said crime has been committed for a specific year.
+The main crimes being analyzed here are assault, auto theft, bike theft, break and enter, robbery, theft over, theft from motor vehicles, homicide, and shooting & firearm discharges.
 The data set also includes the crime rate per 100,000 population.
 That is, it calculates the crime count per 100,000 population per year.
-It is to be noted that for some areas, some crime data may be listed as "NA".
 
-Any other types of crime are not represented in this data set and are deemed to be outside the scope of this paper.
-This data set was last freshed on January 14, 2024 and was captured for this paper on January 16, 2024.
-I conducted the first step of cleaning the data to omit any NA values, simplifying the names of columns, eliminating any additional columns and re-factoring the data (see @tbl-crimedata).
+Other types of crime are not represented in this data set and are deemed to be outside the scope of this paper.
+This data set was last refreshed on January 14, 2024, and was captured for this paper on January 16, 2024.
+I conducted the first step of cleaning the data to omit any NA values, simplifying the names of columns, eliminating any additional columns, and re-factoring the data (see @tbl-crimedata).
 
@@
 
-Further breaking down this data allows us to focus on the trends of crime rates over a specified period and determining the percentage change in the average crime rate to give us an insight into how crime patterns have evolved.
+Further breaking down this data allows us to focus on the trends of crime rates over a specified period and determine the percentage change in the average crime rate to give us an insight into how crime patterns have evolved.
 
-### Calcuating the Percentage Change in Average Crime Rate
+### Calculating the Percentage Change in Average Crime Rate
 
@@
 @tbl-percentagecrimerates20142023 shows the percentage change for each crime when comparing the years 2014 and 2023.
-Notably, we can see that both auto theft and homicide has increased over 210% while crimes such as robbery and theft from motor vehicle have both decreased ~24% and ~21% respectively.
+Notably, we can see that both auto theft and homicide have increased by over 210% while crimes such as robbery and theft from motor vehicle have both decreased by ~24% and ~21% respectively.
 @tbl-percentagecrimerates20222023 shows how homicide is now up 180% while robbery is now up 11% with theft from motor vehicle only down 7%.
@@
 
-Overall, the data reveals that there is a significant upward trend of crime rates, more particularly in homicide and auto theft rates.
+Overall, the data reveals that there is a significant upward trend in crime rates, more particularly in homicide and auto theft rates.
 The data indicates an over 210% increase in homicide rates with a more notable spike of 180% in the last year.
@@
 These findings are in alignment with recent reports from various news outlets that suggest a growing trend in criminal activity.
-The [National Post](https://nationalpost.com/news/canada/violent-crime-rose-again-in-2022-murder-rate-highest-in-30-years-statcan) stating 'murder rate highest in 30 years' [@citearticle3], [CP24](https://www.cp24.com/news/toronto-saw-15-per-cent-increase-in-its-violent-crime-index-in-2022-more-than-three-times-the-jump-seen-nationally-statscan-1.6498269?cache=yesclipId10406200text%2Fhtml%3Bcharset%3Dutf-80404%2F7.258454%2F7.632243%2F7.243626%2F7.243626%2F7.243626) claiming that 'Toronto saw 15 per cent increase in its violent crime index in 2022' [@citearticle1], or the [CBC](https://www.cbc.ca/news/canada/toronto/auto-theft-stories-toronto-2023-1.7060776#:~:text=In%20Toronto%20proper%2C%20vehicle%20thefts,jump%20in%202022%20from%202021.) saying that 'vehicle thefts were up some 25.4 per cent over 2022' [@citearticle2].
+The [National Post](https://nationalpost.com/news/canada/violent-crime-rose-again-in-2022-murder-rate-highest-in-30-years-statcan) stating 'murder rate highest in 30 years' [@citearticle3], [CP24](https://www.cp24.com/news/toronto-saw-15-per-cent-increase-in-its-violent-crime-index-in-2022-more-than-three-times-the-jump-seen-nationally-statscan-1.6498269?cache=yesclipId10406200text%2Fhtml%3Bcharset%3Dutf-80404%2F7.258454%2F7.632243%2F7.243626%2F7.243626%2F7.243626) claiming that 'Toronto saw 15 percent increase in its violent crime index in 2022' [@citearticle1], or the [CBC](https://www.cbc.ca/news/canada/toronto/auto-theft-stories-toronto-2023-1.7060776#:~:text=In%20Toronto%20proper%2C%20vehicle%20thefts,jump%20in%202022%20from%202021.) saying that 'vehicle thefts were up some 25.4 percent over 2022' [@citearticle2].
+An interesting caveat in the data is that there is a noticeable decrease in crime rates when looking from 2020-2021.
+Global events like COVID-19 could have influenced these trends.
 
@@
 What would someone's motive be for committing a crime in the first place?
-This could be down to numerous different factors.
+While the percentage in crime rates has increased over the years, correlation doesn't necessarily mean causation as the reason for these percentage increases could be down to numerous different factors.
 Some factors that could be contributing to this trend could be socio-economic factors.
-For example income inequality, unemployment, high costs of living, etc.
-which are major prevailing problems that all Canadians, not just specifically people living in Toronto, in the last few years has experienced to some degree.
+For example, income inequality, unemployment, high costs of living, etc.
+are major prevailing problems that all Canadians, not just specifically people living in Toronto, in the last few years have experienced to some degree.
 Another factor could be the societal attitude towards crime.
-If someone were to commit a crime, there is obviously an incentive to do so.
-In the case of auto theft, thieves can get away with stealing an entire car and then selling it over-seas.
+If someone were to commit a crime, there is an incentive to do so.
+In the case of auto theft, thieves can get away with stealing an entire car and then selling it overseas.
 The point to highlight here is they can get away with doing said crime.
-If the incentive to commit the crime far out-weighs any potential punishment for said crime, then it gives criminals the encouragement to do the crime and more importantly repeat doing so.
+If the incentive to commit the crime far outweighs any potential punishment for said crime, then it gives criminals the encouragement to do the crime and more importantly repeat doing so.
+
 This increase in crime rate calls for an urgent review of our current crime prevention and law enforcement strategies.
+This can include giving better funding to our law enforcement and crime prevention programs or making major changes to the policies addressing socio-economic issues.
 
-There are several limitations to this analysis 
+There are several limitations to this analysis and it is important to acknowledge these limitations of our analysis.
+A major limitation in our analysis is that this data only accounts for reported crime.
+While trying to factor in unreported crime is deemed to be beyond the scope of this paper, it is important to highlight that our data set may not capture unreported crime incidents.
+By extension, the argument could also be made that the 'increase' in crime rate could be because of a more efficient crime reporting system or changes in the legal definitions of certain crimes.
+Another limitation is that the raw data set listed some crime rates as 'NA'.
+This can cause our data to be skewed in certain places and could potentially mislead our findings.
+
 # Conclusion
 
-This paper investigated the
+This paper investigated the crime rates over the years in Toronto taking into account the 158 different neighborhoods in Toronto.
+My results reveal that there was an increase in crime rate overall over the years, with crimes such as homicide and auto theft having the more notable and concerning increases.
+However, these results don't necessarily imply causation, but rather a high correlation that warrants a deeper investigation.
+Future research should be aimed to explore the factors as to why this increase is happening and why these trends are occurring.
+Whether it be to examine the current effectiveness of various crime prevention strategies or to monitor ongoing crimes to provide a better understanding of the nature of the crime which can be crucial to implementing a timely and effective response.
+Having strong law enforcement is crucial for ensuring the safety of the general public and to make sure new and repeat offenders are dealt with by justice.
 

### 2024-01-24 outputs/paper/paper.qmd
Changed number of decimals to 2 and updated README

@@
 
-head(cleaned_crime_rates_data) |> kable(col.names = c("Crime", "Year", "Crime Rate"))
+head(cleaned_crime_rates_data) |> kable(col.names = c("Crime", "Year", "Crime Rate"), digits = 2)
 ```
@@
 
-head(average_crime_rates_per_year) |> kable(col.names = c("Crime", "Year", "Average Crime Rate"))
+head(average_crime_rates_per_year) |> kable(col.names = c("Crime", "Year", "Average Crime Rate"), digits = 2)
 ```
@@
 
-final_percentage_change_table |> select("Crime", "Percentage Change") |> kable()
+final_percentage_change_table |> select("Crime", "Percentage Change") |> kable(digits = 2)
 ```
@@
 colnames(percentage_change_table) <- c("X", "Crime", "Percentage Change")
-percentage_change_table |> select("Crime", "Percentage Change") |> kable()
+percentage_change_table |> select("Crime", "Percentage Change") |> kable(digits = 2)
 ```

### 2024-01-24 outputs/paper/paper.qmd
updated paper and scripts

@@
 date-format: long
-abstract: "Current mainstream news outlets are reporting a significant increase in crime rates, a trend that from our data analysis supports.
+abstract: "Current mainstream news outlets are reporting a significant increase in crime rates, a trend that our data analysis supports.
 Our findings indicate a dramatic increase in crimes such as homicide and auto theft, with them increasing over 210% from 2014.
@@
 This paper is structured into the following sections: Data, Results, Discussion, and Conclusion.
-In the data section I discuss the nature of these crime rates obtained from the City of Toronto's `opendatatoronto` [@citeOpenDataToronto] library, and the steps I took to clean and analyze the data.
-The result section highlights the trends found in the analysis process.
-The discussion section provides further evaluation of the trends and serves to provide insight as to the potential reasoning behind these crime trends and possible areas that require more attention.
-Lastly, the conclusion section is to summarize the main findings of this paper.
+In @sec-data, I discuss the nature of these crime rates obtained from the City of Toronto's `opendatatoronto` [@citeOpenDataToronto] library, and the steps I took to clean and analyze the data.
+@sec-results highlights the trends found in the analysis process.
+@sec-discussion provides further evaluation of the trends and serves to provide insight as to the potential reasoning behind these crime trends and possible areas that require more attention.
+Lastly, @sec-conclusion is to summarize the main findings of this paper.
 
@@
 
-# Results
+# Results {#sec-results}
 
@@
 
-# Creating the line graph for average crime rates
+# Changing the names on the graph legend
+average_crime_rates_per_year$crime <- factor(average_crime_rates_per_year$crime,
+                                             levels = c("assault", "autotheft", "biketheft", "breakenter", "homicide", "robbery", "shooting", "theftfrommv", "theftover"),
+                                             labels = c("Assault", "Auto Theft", "Bike Theft", "Break and Enter", "Homicide", "Robbery", "Shooting", "Theft from Motor Vehicle", "Theft over"))
+
+# Creating the line graph with updated legend names
 ggplot(average_crime_rates_per_year, aes(x = year, y = average_rate, color = crime)) +
@@
 
+# Changing the names on the graph legend
+percentage_change_data$crime <- factor(percentage_change_data$crime,
+                                             levels = c("assault", "autotheft", "biketheft", "breakenter", "homicide", "robbery", "shooting", "theftfrommv", "theftover"),
+                                             labels = c("Assault", "Auto Theft", "Bike Theft", "Break and Enter", "Homicide", "Robbery", "Shooting", "Theft from Motor Vehicle", "Theft over"))
+
 # Creating the line graph for percentage changes
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 
@@
 
-# Conclusion
+# Conclusion {#sec-conclusion}
 

### 2024-01-24 scripts/00-simulate_data.R

@@ -5,9 +5,6 @@
 # Date: 16 January 2024
 # Contact: adrian.ly@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
 
 #### Workspace setup ####
 library(knitr)
